---
title: "Data301 Assignment 4"
author: "Cory Stalker"
date: "2026-09-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Library Loading

```{r setup, include=FALSE}
library(tidyverse)

```

#--------------------------------

# Question 1

## Part 1 (10%): Read and summarise the analysis of alternatives made by Wicklin

(2017). This article gives you a link to the article that sparked the
discussion (Hyndman & Fan, 1996), and we strongly recommend you also
read it. Which type of order statistics would you use and why? Which is
the default type of order statistics your version of R/Python uses?

Answer: Hyndman & Fan showed that statistical software computes sample
quantatiles in multiple different ways (9). Three versions of quantile
computation are based on rounding and the other six are based off linear
interpolation (a way to estimate an unknown value that lies between two
known data points, assuming a linear relationship). SAS uses type 2 and
R uses type 7.

Wicklin writes a general formula to cover all quantile estimates, q =
(1-$\lambda$) x[j] + $\lambda$ x[j+1], where j depends on the values of
p, N and m (j = Np + m). Wicklin compares the definitions visually on a
10-observation dataset and on 100 quantiles from a uniform
distributions, also comparing the R and SAS defaults.

Wicklin notes that the rounding-based types return an observed value or
the average of two neighbours, while the interpolation types may return
any value in the data's range. For the dataset {0,1,1,1,2,2,2,4,5,8},
the two defaults disagree on this dataset. For the first quartile, both
give 1. However, for the third quartile, SAS (or type 2) gives 4, while
R (or type 7) gives \~3.5. Applying these results to:
$\hat{\lambda}_{IQR}$(X) =
$\frac{ln3}{q_{\frac{3}{4}}(X) - q_{\frac{1}{4}}(X)}$ The estimate of
$\lambda$ is from the following: Type 2: $\frac{ln(3)}{4-1}$ =
$\frac{1.098612}{3}$ = \~0.3662 (4dp) Type 7: $\frac{ln(3)}{3.5-1}$ =
$\frac{1.098612}{2.5}$ = \~0.4394 (4dp) A difference of of \~0.0732
(4dp) or \~18.17% is observed between the two estimates of $\lambda$.
Although Wicklin states that with n = 100 \~N(0, 1) the difference are
nearly negligible.

The experiments Wicklin run show that the choice of quantile computation
type matter for smaller samples. However for larger scale data, the
differences are negligible. He does not recommend an exact type of
quantile computation, but mentions that "most data analysts just accept
the default method for whatever software they are using"[1].

I would most likely use type 8, this is because it is recommended by
Hyndman and Fan (1996) [2]. Type 8 can be defined independently of the
underlying distribution. Wicklin showed that the definitions differ
mostly for small samples, so the choice matters when n is small. My
current version of R ("R version 4.4.2 (2024-10-31)") uses type 7 as the
default type of order statistics.

## Part 2 (20%):

You suspect that $\hat{\lambda}_{IQR}$ is biased, so derive and
implement a first-order bootstrap bias corrected estimator based on
$\hat{\lambda}_{IQR}$. Denote it $\hat{lambda}\_{IQR}$. You may fix the
number of bootstrap replications in $𝐵$ = 200 for every input sample.

Answer:

let $\frac{ln3}{\hat{q}_{\frac{3}{4}}(X) - \hat{q}_{\frac{1}{4}}(X)}$
with quartiles be computed by type 8. For an estimator $\hat{\theta}$ of
$\theta$, the bias is $B(\hat{\theta}) = E(\hat{\theta}) - \theta$, here
$\theta = \lambda$ and $\hat{\theta} = \hat{\lambda}_{IQR}$. The bias
$B(\hat{\lambda}_{IQR}) = E(\hat{\lambda}_{IQR}) - \lambda$ (Monte Carlo
lecture, slide 21/34 (or pdf page 43), under the
law$D(\lambda) = EXP(\lambda)$ of $X = (X_1,X_2...X_N)$. This can't be
computed because $D(\lambda)\$$ depends on the unknown,so I will
estimate it by the bootstrap.I resample the observed x with replacement
B = 200 times and calculate $\hat{\lambda}\_{IQR}((x)(b))$ (Monte Carlo
slide 22 or pdf page 44), giving:

$\hat{B}(\hat{\lambda}_{{IQR}}) = B^{-1} \sum_{b=1}^{B} \hat{\lambda}_{{IQR}}\bigl(x^{*(b)}\bigr) - \hat{\lambda}_{{IQR}}(x).$

subtracting it gives:

$\tilde{\lambda}= \hat{\lambda}_{{IQR}}(x) - \hat{B}(\hat{\lambda}_{{IQR}}) = 2\,\hat{\lambda}_{{IQR}}(x) - B^{-1} \sum_{b=1}^{B} \hat{\lambda}_{{IQR}}\bigl(x^{*(b)}\bigr).$

```{r}
lambda_iqr = function(x){ 
  log(3) / diff(quantile(x, c(.25, .75), type = 8, names = FALSE))
}

lambda_tilde = function(x, B =200){
  lamda_hat = lambda_iqr(x)
  lambda_bootstrap = replicate(B, lambda_iqr(sample(x, replace = TRUE)))
  lambda_bootstrap = lambda_bootstrap[is.finite(lambda_bootstrap)]
  2 * lamda_hat - mean(lambda_bootstrap)
}

set.seed(301)
# sample 30 random numbers from an exponential distribution
x = rexp(30, rate = 1)
c(lam_iqr = lambda_iqr(x), lam_tilde = lambda_tilde(x))
```

# References:

[1]
<https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html>
\n
[2]<https://www.amherst.edu/media/view/129116/original/Sample+Quantiles.pdf>
