---
title: "Data301 Assignment 4"
author: "Cory Stalker"
date: "2026-09-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Library Loading

```{r setup, include=FALSE}
library(tidyverse)
library(fpp2)
library(ggplot2)
library(lubridate)
library(pander)
```

#--------------------------------

# Question 1

## Part 1 (10%): Read and summarise the analysis of alternatives made by Wicklin

(2017). This article gives you a link to the article that sparked the
discussion (Hyndman & Fan, 1996), and we strongly recommend you also
read it. Which type of order statistics would you use and why? Which is
the default type of order statistics your version of R/Python uses?

Answer: Hyndman & Fan showed that statistical software computes sample
quantatiles in multiple different ways (9). Three versions of quantile
computation are based on rounding and the other six are based off linear
interpolation (a way to estimate an unknown value that lies between two
known data points, assuming a linear relationship). SAS uses type 2 and
R uses type 7.

Wicklin writes a general formula to cover all quantile estimates, q =
(1-$\lambda$) x[j] + $\lambda$ x[j+1], where j depends on the values of
p, N and m (j = Np + m). Wicklin compares the definitions visually on a
10-observation dataset and on 100 quantiles from a uniform
distributions, also comparing the R and SAS defaults.

Wicklin notes that the rounding-based types return an observed value or
the average of two neighbours, while the interpolation types may return
any value in the data's range. For the dataset {0,1,1,1,2,2,2,4,5,8},
the two defaults disagree on this dataset. For the first quartile, both
give 1. However, for the third quartile, SAS (or type 2) gives 4, while
R (or type 7) gives \~3.5. Applying these results to:
$\hat{\lambda}_{IQR}$(X) =
$\frac{ln3}{q_{\frac{3}{4}}(X) - q_{\frac{1}{4}}(X)}$ The estimate of
$\lambda$ is from the following: Type 2: $\frac{ln(3)}{4-1}$ =
$\frac{1.098612}{3}$ = \~0.3662 (4dp) Type 7: $\frac{ln(3)}{3.5-1}$ =
$\frac{1.098612}{2.5}$ = \~0.4394 (4dp) A difference of of \~0.0732
(4dp) or \~18.17% is observed between the two estimates of $\lambda$.
Although Wicklin states that with n = 100 \~N(0, 1) the difference are
nearly negligible.

The experiments Wicklin run show that the choice of quantile computation
type matter for smaller samples. However for larger scale data, the
differences are negligible. He does not recommend an exact type of
quantile computation, but mentions that "most data analysts just accept
the default method for whatever software they are using"[1].

I would most likely use type 8, this is because it is recommended by
Hyndman and Fan (1996) [2]. Type 8 can be defined independently of the
underlying distribution. Wicklin showed that the definitions differ
mostly for small samples, so the choice matters when n is small. My
current version of R ("R version 4.4.2 (2024-10-31)") uses type 7 as the
default type of order statistics.

## Part 2 (20%):

You suspect that $\hat{\lambda}_{IQR}$ is biased, so derive and
implement a first-order bootstrap bias corrected estimator based on
$\hat{\lambda}_{IQR}$. Denote it $\hat{lambda}\_{IQR}$. You may fix the
number of bootstrap replications in $𝐵$ = 200 for every input sample.

Answer:

let $\frac{ln3}{\hat{q}_{\frac{3}{4}}(X) - \hat{q}_{\frac{1}{4}}(X)}$
with quartiles be computed by type 8. For an estimator $\hat{\theta}$ of
$\theta$, the bias is $B(\hat{\theta}) = E(\hat{\theta}) - \theta$, here
$\theta = \lambda$ and $\hat{\theta} = \hat{\lambda}_{IQR}$. The bias
$B(\hat{\lambda}_{IQR}) = E(\hat{\lambda}_{IQR}) - \lambda$ (Monte Carlo
lecture, slide 21/34 (or pdf page 43), under the
law$D(\lambda) = EXP(\lambda)$ of $X = (X_1,X_2...X_N)$. This can't be
computed because $D(\lambda)\$$ depends on the unknown,so I will
estimate it by the bootstrap.I resample the observed x with replacement
B = 200 times and calculate $\hat{\lambda}\_{IQR}((x)(b))$ (Monte Carlo
slide 22 or pdf page 44 and Monte Carlo lecture recording), giving:

$\hat{B}(\hat{\lambda}_{{IQR}}) = B^{-1} \sum_{b=1}^{B} \hat{\lambda}_{{IQR}}\bigl(x^{*(b)}\bigr) - \hat{\lambda}_{{IQR}}(x).$

subtracting it gives:

$\tilde{\lambda}= \hat{\lambda}_{{IQR}}(x) - \hat{B}(\hat{\lambda}_{{IQR}}) = 2\,\hat{\lambda}_{{IQR}}(x) - B^{-1} \sum_{b=1}^{B} \hat{\lambda}_{{IQR}}\bigl(x^{*(b)}\bigr).$

```{r}
lambda_iqr = function(x){ 
  log(3) / diff(quantile(x, c(.25, .75), type = 8, names = FALSE))
}

lambda_tilde = function(x, B =200){
  lamda_hat = lambda_iqr(x)
  lambda_bootstrap = replicate(B, lambda_iqr(sample(x, replace = TRUE)))
  lambda_bootstrap = lambda_bootstrap[is.finite(lambda_bootstrap)]
  2 * lamda_hat - mean(lambda_bootstrap)
}

set.seed(301)
# sample 30 random numbers from an exponential distribution
x = rexp(30, rate = 1)
c(lam_iqr = lambda_iqr(x), lam_tilde = lambda_tilde(x))
```

## Part 3 (20%): Make a Monte Carlo experiment comparing these three estimators. Use as parameter space $\Lambda \times$ 𝑁, where $\lambda \in\Lambda$ = {.1, 1, 10} and $n \in$ 𝑁 = {5, 30, 100}. Use the mean square error as the sole metric for your analysis. Report your results. Justify your choice(s) of the number of replications for each $\lambda$ and each $n$. You may want to check the book by Robert & Casella (2009); it is available in our library.

```{r}
lambda_ml = function(x) 1 / mean(x)

lambda = c(.1, 1,10)
n = c(5,30,100)
rmax = 1000

results = data.frame(lambda = numeric(0), n =numeric(0)
                     ,mse_max_likelihood = numeric(0), mse_iqr = numeric(0), mse_tilde = numeric(0))

for (l in lambda){
  for (num in n){
    squared_err_ml = squared_err_iqr = squared_err_tilde = numeric(rmax)
    for (r in 1:rmax){
      x =rexp(num, rate = l)
      squared_err_ml[r] = (lambda_ml(x) - l)^2
      squared_err_iqr[r] = (lambda_iqr(x) - l)^2
      squared_err_tilde[r] = (lambda_tilde(x) - l)^2
    }
    
    results = rbind(results, data.frame(
      l = l, n= num,
      mse_max_likelihood = mean(squared_err_ml),
      mse_iqr = mean(squared_err_iqr, na.rm = TRUE),
      mse_tilde = mean(squared_err_tilde, na.rm = TRUE)
    ))
  }
}
results
```

```{r}
lambda <- 10; n_val <- 5; rmax <- 5000
se <- numeric(rmax)
for (r in 1:rmax) {
  x <- rexp(n_val, rate = lambda)
  se[r] <- (lambda_tilde(x) - lambda)^2
}
summary(se)
sort(se, decreasing = TRUE)[1:10]   # look at the biggest outliers
```

#--------------------------------

# Question 2

## Part 1: What is the frequency of the series? How can you determine this from the data and/or the time-series plot?

```{r, fig.width= 12, fig.height = 6}
data(ausbeer)
ausbeer

cat("Frequency:",frequency(ausbeer), "\n")
cat("Start:",start(ausbeer),"\n")
cat("End:",end(ausbeer))
summary_df = data.frame(
  Frequency = frequency(ausbeer),
  Start = paste(start(ausbeer)),
  End = paste(end(ausbeer))
)


autoplot(ausbeer) +
  labs(title = "Australian Quarterly Beer Production", x = "Year", y = "Megalitres") +
  geom_point(size = 1.5, color = "red")+
  scale_x_continuous(breaks = seq(1956, 2010, by =1))+
  theme(axis.text.x = element_text(angle = 45, hjust =1))
  
```

The Frequency of the series can be found by using the frequency()
function on the data, this returns 4 observations per (quarterly data).
You can also visualise this in two ways, the data table has 4 columns
for each year, and the time series plot shows 4 points within each year.

## Part 2: Describe the overall behaviour of the series. Is there evidence of a long-term trend?

```{r}
autoplot(ausbeer) +
  labs(title = "Australian Quarterly Beer Production", x = "Year", y = "Megalitres")
```

The time series graph does not show a single, consistent or monotonic
trend across the datasets whole period. Production rises steadily from
around 250 megalitres to \~550 megalitres by the mid 1970s. It then
plateaus, oscillating without any clear patterns to \~1990. From the
1990s onwards, production declined gradually to around \~450 - \~500
megalitres by 2010.

## Part 3: Is there evidence of seasonality? If so, describe the seasonal pattern.

Yes there is a clear pattern of seasonality, this is visible in nearly
every year. Q4 is consistently the peak quarter across the entire series
(1956-2009), with production rising sharply toward the end of each year.
Production sharply rises towards the end of year which makes sense as
this is summer and Christmas time in Australia and New Zealand, where
beer consumption would be expected to rise.

The low point is usually Q2 in most years, however in 7 years this
differs (1975, 1978, 1984m 1990, 1993, 1996, and 2002). In these years
Q3 is equal to or lower than Q2 instead. So while the overall trend is a
rise in Q4 followed by a fall is stable, the yearly-quarter of the
minimum varies throughout the years.

## Part 4: Does the variability appear constant over time? Explain what you see in the plot and whether a transformation might be appropriate.

```{r}
beer_matrix = matrix(window(ausbeer, start = c(1956,1),end = c(2009, 4)), ncol = 4, byrow = TRUE)

rownames(beer_matrix) = 1956:2009
yearly_variance = apply(beer_matrix, 1 ,var)
yearly_standard_dev = apply(beer_matrix, 1, sd)
yearly_range = apply(beer_matrix, 1, function(row) max(row) - min(row))

range_df = data.frame(
  year = 1956:2009,
  range = yearly_range
)

pander(range_df, title = "Yearly Range of Beer Production")

variability_df = data.frame(
  year = 1956:2009,
  sd = yearly_standard_dev
)
ggplot(variability_df, aes(x = year, y= sd))+
  geom_line() +
  labs(title = "Variability of quarterly production by year",
       x = "Year", y="Within-Year standard deviation (megalitres)")
```

From both the range table and the Variability of quarterly production by
year plot, it is obvious that variance is not constant over time. The
within-year standard deviation of quarterly production for each year
shows as

# References:

[1]
<https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html>
\n
[2]<https://www.amherst.edu/media/view/129116/original/Sample+Quantiles.pdf>
